Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time

### 2025-02-14 ms-3/ms.tex
Update ms.tex

% Redefine the \figurename to make it bold
\renewcommand{\figurename}{\textbf{Figure}}

@@

% Set citation style
\setcitestyle{numbers}

@@

\clearpage
\begin{figure}[!ht]
\includegraphics[width=1\textwidth]{fig1}
\caption{
Detecting dispersion changes in case count time series. 
a: Weekly incidence of COVID-19 in the United States, with time measured in weeks since January 4, 2020, showing an example of a randomly-selected 16 week period used as an incidence trend when in simulation-based validation of the LRT test (red). 
b: Cases in in one county (Douglas County, Nebraska) over the sample time period with estimated incidence trend (red) and estimated dispersion values on either side of the midpoint. 
c: Estimated $\theta$ versus true $\theta$ in simulation studies combining a randomly-selected section of the national incidence curve with a random population size and set of dispersion values. Estimated values outside of tolerance plotted in purple (close to Poisson) and blue (close to collapsing to zero). 
d: Statistical power of the LRT test.}\label{fig1}
\end{figure}
\clearpage

Applying the method to COVID-19 cases in US counties enabled investigation of changes in dispersion in relation to both observed trends and model expectations based on case counts (Fig~\ref{fig2}).  
Periods of increased case count variability corresponded with decreases in $\theta$, indicating that dispersion was dynamic. 
Changes in dispersion exhibited both expected and unexpected patterns of variation relative to standard theory.  
In some instances, $\theta$ varied inversely with incidence, consistent with standard epidemic theory, while in other periods, deviations from this expectation occurred, potentially signaling shifts in underlying transmission dynamics.  
Notably, significant changes in the dispersion parameter were observed during major epidemic transitions.
For example, during the beginning of 2022, and at the end of the time series, when the pandemic was transitioning toward endemicity as the landscape of susceptibility was evolving~\citep{lavine2021immunological}.  
The landscape of susceptibility was evolving as a larger proportion of cases involved reinfections.  
These findings underscore the complex behavior of the dispersion parameter, which not only varied with changes in case count regimes but also revealed departures from the model expectations described by Eq.~1, which are consistent with changes in the underlying drivers of transmission.

\clearpage
\begin{figure}[!ht]
\includegraphics[width=1\textwidth]{fig2}
\caption{Dispersion analysis of weekly COVID-19 case data for Jefferson County, Alabama. 
Results for all counties are shown in Figure 3  %.~\ref{fig3}.  
a: Weekly reported COVID-19 incidence.  
b: Estimated dispersion parameter ($\theta$) over time.  
c: Comparison of estimated dispersion (gray) with predicted values from the standard model $\theta_{t+1} = C_t / \rho_t$, where $C_t$ is reported cases and $\rho_t$ is the reporting rate. Predictions are shown for fixed $\rho_t = 0.1$ (black) and $\rho_t = 0.9$ (blue), chosen to encompass the range of $\theta$ expected under variable $\rho$.  
d: Likelihood ratio test (LRT) statistic over time. Statistically significant changes in dispersion (red) correspond to p-values below the Bonferroni-corrected 5\% threshold of a chi-square distribution with one degree of freedom.  
}\label{fig2}
\end{figure}
\clearpage

@@

\clearpage
\begin{figure}[!ht]
\includegraphics[width=1\textwidth]{fig3}
\caption{Incidence and dispersion between Jan 4, 2020 and March 18, 2023, in large counties in the US.
a: Mean COVID-19 cases of the 144 US counties over time. 
b: Mean $\log_{10}\theta$ of the 144 US counties over time.  
c: $\log_{10}(cases)$ over time for each of the 144 counties.
d: $\log_{10}\theta$ over time for each of the 144 counties.
e: Expected value of $\log_{10}\theta_0$ under the null model, assuming a reporting rate of 0.5.
f: LRT p-values over time for each location.}\label{fig3}
\end{figure}
\clearpage

>>>>

@@

% Remove parentheses around equation numbers
\renewcommand{\eqref}[1]{Eq.~\ref{#1}}

% Change figure references to "Fig. X" instead of "Figure X"
\renewcommand{\figurename}{Fig.}

% Set citation style
\setcitestyle{square, numbers}

@@

\clearpage
\begin{figure}[!ht]
\includegraphics[width=1\textwidth]{fig1}
\caption{{\bf Fig 1. Detecting dispersion changes in case count time series.}
a: Weekly incidence of COVID-19 in the United States, with time measured in weeks since January 4, 2020, showing an example of a randomly-selected 16 week period used as an incidence trend when in simulation-based validation of the LRT test (red). 
b: Cases in in one county (Douglas County, Nebraska) over the sample time period with estimated incidence trend (red) and estimated dispersion values on either side of the midpoint. 
c: Estimated $\theta$ versus true $\theta$ in simulation studies combining a randomly-selected section of the national incidence curve with a random population size and set of dispersion values. Estimated values outside of tolerance plotted in purple (close to Poisson) and blue (close to collapsing to zero). 
d: Statistical power of the LRT test.}\label{fig1}
\end{figure}
\clearpage

Applying the method to COVID-19 cases in US counties enabled investigation of changes in dispersion in relation to both observed trends and model expectations based on case counts (Fig~\ref{fig2}).  
Periods of increased case count variability corresponded with decreases in $\theta$, indicating that dispersion was dynamic. 
Changes in dispersion exhibited both expected and unexpected patterns of variation relative to standard theory.  
In some instances, $\theta$ varied inversely with incidence, consistent with standard epidemic theory, while in other periods, deviations from this expectation occurred, potentially signaling shifts in underlying transmission dynamics.  
Notably, significant changes in the dispersion parameter were observed during major epidemic transitions.
For example, during the beginning of 2022, and at the end of the time series, when the pandemic was transitioning toward endemicity as the landscape of susceptibility was evolving~\citep{lavine2021immunological}.  
The landscape of susceptibility was evolving as a larger proportion of cases involved reinfections.  
These findings underscore the complex behavior of the dispersion parameter, which not only varied with changes in case count regimes but also revealed departures from the model expectations described by Eq ~1, which are consistent with changes in the underlying drivers of transmission.

\clearpage
\begin{figure}[!ht]
\includegraphics[width=1\textwidth]{fig2}
\caption{{\bf Fig 2. Dispersion analysis of weekly COVID-19 case data for Jefferson County, Alabama.} 
Results for all counties are shown in Figure 3  %.~\ref{fig3}.  
a: Weekly reported COVID-19 incidence.  
b: Estimated dispersion parameter ($\theta$) over time.  
c: Comparison of estimated dispersion (gray) with predicted values from the standard model $\theta_{t+1} = C_t / \rho_t$, where $C_t$ is reported cases and $\rho_t$ is the reporting rate. Predictions are shown for fixed $\rho_t = 0.1$ (black) and $\rho_t = 0.9$ (blue), chosen to encompass the range of $\theta$ expected under variable $\rho$.  
d: Likelihood ratio test (LRT) statistic over time. Statistically significant changes in dispersion (red) correspond to p-values below the Bonferroni-corrected 5\% threshold of a chi-square distribution with one degree of freedom.  
}\label{fig2}
\end{figure}
\clearpage

@@

\clearpage
\begin{figure}[!ht]
\includegraphics[width=1\textwidth]{fig3}
\caption{{\bf Fig 3. Incidence and dispersion between Jan 4, 2020 and March 18, 2023, in large counties in the US.}
a: Mean COVID-19 cases of the 144 US counties over time. 
b: Mean $\log_{10}\theta$ of the 144 US counties over time.  
c: $\log_{10}(cases)$ over time for each of the 144 counties.
d: $\log_{10}\theta$ over time for each of the 144 counties.
e: Expected value of $\log_{10}\theta_0$ under the null model, assuming a reporting rate of 0.5.
f: LRT p-values over time for each location.}\label{fig3}
\end{figure}
\clearpage
